---
title: "Exploratory Analysis of Seattle Collisions"
author: "Leah Erb"
---

This report explores **collision** report data from **Seattle, Washington, USA**.

# The Dataset
The [Seattle Collisions](https://data-seattlecitygis.opendata.arcgis.com/datasets/collisions) dataset is a compilation of over 200,000 collision reports created by Seattle Police Department (SPD) that were then recorded by Seattle Department of Transportation (SDOT), between the years 2004 and 2018.


```{css echo=FALSE, message=FALSE, warning=FALSE, CSS}
/* Define a margin before header elements */
h1  {
  margin-top: 52px;
}
h2, h3, h4  {
  margin-top: 48px;
}
/* Define a margin after every first p elements */
p:first-of-type {
  margin-bottom: 14px;
}
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE, Libraries}
# libraries
library(ggplot2)
library(ggpubr)     # ggarrange
library(scales)
library(GGally)     # ggcorr
library(dplyr)
library(tidyverse)
library(knitr)
library(psych)
library(reshape2)  
library(cowplot)
library(gridExtra)

# global settings
knitr::opts_chunk$set(comment=NA, fig.width=9, fig.height=6)

# ggplot layers
mytheme_minimal <- theme_minimal() +
    theme(plot.title = element_text(size = 16, hjust = 0),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 11),
          axis.title.x = element_text(hjust = 0),
          legend.position = "none"
          ) 

mylayer_color <- scale_color_brewer(palette = 'Dark2', direction = 1)  
mylayer_fillcolor <- scale_fill_brewer(palette = 'Dark2', direction = 1)
mylayer_x90 <- theme(axis.text.x = element_text(angle = 90, hjust = 1))
mylayer_hour_breaks <- scale_x_datetime(labels = date_format("%H:%M"), date_breaks = ('2 hours')) 
mylayer_long_xlabel <- theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
mylayer_count_above <- geom_text(stat='count', aes(label = ..count..), vjust = -.2)
```

# Load data
```{r Load_data}
datafile = '../data/collisions_clean.rda'
load(file = datafile)

# Create df_trim_yrs to contain only complete year's worth of data
# TODO: include logic, remove hardcode
df_full_yrs <- subset(df, year != '2019' & year != '2003')
df$hour <- as.integer(df$hour)
```

# Explore

## When

### 1. Collision count by year
```{r}
ggplot(data = df_full_yrs %>% count(year), aes(x = year, y = n)) + 
    geom_bar(stat = "identity", aes(fill = n)) + 
    labs(title = 'Collision count by year', x = '', y = '') +
    mytheme_minimal +
    scale_fill_distiller(palette = 'Greys', direction = 1) +
    scale_x_discrete(position = 'top') +
    scale_y_continuous(labels = scales::comma) 
```

**Observations: **
There appears to be a 10-is year trend: 5 years down, 5 years up, then going back down.

The first peak, at year 2005, was higher than the second peak 10 years later, at year 2015.

I had expected to see a steady increase in Seattle collisions, mirroring population growth. Perhaps I'm wrong about the steady increase in population, or wrong that the number of **reported** collisions have a strong correlation with the population. #TODO: import population data and compare rate changes with collision rates by year.

### 2. Collision count by hour
```{r warning=FALSE}
 df %>% 
    filter(!is.na(hour)) %>%
    count(hour)

ggplot(data = df %>% 
           filter(!is.na(hour)) %>%
           count(hour), 
       aes(x = hour, y = n, group = 1 )) + 
    geom_line() + 
    labs(title = 'Collision Count by Hour', x = '', y = '') +
    mytheme_minimal  +
    scale_y_continuous(labels = scales::comma) 
```

### 2b. Collision count by hour, Radial version

```{r}

create_radial_plot <- function(data, col_name, title, subtitle) {
    
    col_name <- enquo(col_name)
    
    # define inner and outer circle sizes (y)
    max_col <- as.integer(
        data %>%
            group_by(!!col_name) %>%
            summarise(n = n()) %>%
            top_n(1) %>%
            .$n)
    # size the hole for the middle of radial at 5% of the radius
    donut_hole <- max_col * .05  
    # define y values for reference circles at 1/3 intervals
    b1 <- max_col * .33
    b2 <- max_col * .66
    b3 <- max_col * .99
 
    p <- ggplot() +
        # make radial
        coord_polar(start=-48/360) +
        # add circle grid lines at defined breaks
        geom_hline(yintercept = c(0, b1, b2, b3), colour = my_grid_grey, size = 0.2) +
        # draw radius grid manually to avoid extending beyond outer circle
        geom_vline(xintercept = seq(0, 23), colour = my_grid_grey, size = 0.2) +
        # draw data 'slices'
        geom_bar(data = data, aes(x = hour), width=1, color='white', size=0.1)  +
        # mark hours on the 24-hour clock
        scale_x_continuous(breaks = seq(0,23), 
                           labels = c('12 am', paste(1:11,'am'), '12 pm', paste(13:23-12, 'pm'))) +
        # make donut hole
        scale_y_continuous(limits = c(-donut_hole, max_col + donut_hole)) + 
        # add plot titles
        labs(title = title,
             subtitle = subtitle,
             caption = 'Data source: https://data.seattle.gov  (2004 - 2018)') +
        # define theme options
        theme_minimal() +
        theme(
            plot.title = element_text(size = 24, face = 'bold', color = '#333333'),
            plot.subtitle = element_text(size = 12, color = '#333333'),
            plot.caption = element_text(size = 7), #, hjust = 1.8),
            # bold-ize 6 am/pm and 12 am/pm
            axis.text.x = element_text(size = 12,
                                       face = c('bold', 'plain', 'plain', 'plain', 'plain', 'plain',
                                                'bold', 'plain', 'plain', 'plain', 'plain', 'plain',
                                                'bold', 'plain', 'plain', 'plain', 'plain', 'plain',
                                                'bold', 'plain', 'plain', 'plain', 'plain', 'plain')),
            # remove axis labels
            axis.title = element_blank(),
            # remove y axis ticks and text
            axis.text.y = element_blank() , 
            axis.ticks = element_blank(),
            # remove outermost circle
            panel.grid.major.y = element_blank(),
            # remove 1/2 hour grid lines
            panel.grid.minor = element_blank(),
            # remove hour grid lines, too
            panel.grid.major.x = element_blank()
        )  
 
    
    return (p)
}
```

```{r message=FALSE, warning=FALSE, error=FALSE, color_options}
# grideline grey setting
my_grid_grey <- 'grey90'
# All collisions
create_radial_plot(data = df, 
            col_name = hour,
            title = '', 
            subtitle = 'All collisions')
# DUI
create_radial_plot(data = df[which(df$DUI == 'Y'),], 
            col_name = hour,
            title = '', 
            subtitle = 'DUIs')
```




**Observations: **
Collisions tend to rise during commute hours, starting at the lowest point around 4am, reaching a peak at 5pm. 

#### 2a. Hourly collisions by day of week
```{r warning=FALSE}
weekend <- c('Saturday', 'Sunday')
weekday <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
alldays <- c(weekday, weekend)

# TODO fix y max limit programmatically
plot_hours <- function(day, title) {
    gg <- ggplot(data = df %>%
               filter(!is.na(hour) & weekdays(INCDATE) %in% day) %>%
               count(hour),
           aes(x = hour, y = n, group = 1 )) +
        geom_line() +
        labs(title = paste(title), x = '', y = '') +
        mytheme_minimal  +
        scale_x_discrete(breaks = c(0,4,8,12,16,20,24)) +
        scale_y_continuous(limits = c(0, 2250), 
                           breaks = seq(0, 2100, 700),
                           labels = scales::comma) 
    return (gg)
}

# create list of plots, one for each day of the week
plot_list <- lapply(
    as.list(alldays), function(x){
    plot_hours(x, paste0('- ', x, 's'))
})

#cowplot::plot_grid(plotlist = plot_list, ncol = 3)

ggarrange(plotlist = plot_list, ncol=3, nrow=3) + 
    annotate(geom = 'text', label = 'title', hjust = 0)

```


### By month

### By part of day

### By season

## Who

### Striker, strikee

### Pedestrians, Cyclists

### Parked cars, trains, other

## What

### Collision type

### Property damage, Injuries, Fatalities

## Where

### Address type, Junction type

## Why

### Weather, Road Condition, Lighting

### DUI, Distracted, Speeding

### Traffic volume(?)

## Multivariate

### When and Why

### Who and What

### <etc>

# Summary

||||||| merged common ancestors
=======
---
title: "Exploratory Analysis of Seattle Collisions"
author: "Leah Erb"
---

This report explores **collision** report data from **Seattle, Washington, USA**.

# The Dataset
The [Seattle Collisions](https://data-seattlecitygis.opendata.arcgis.com/datasets/collisions) dataset is a compilation of over 200,000 collision reports created by Seattle Police Department (SPD) that were then recorded by Seattle Department of Transportation (SDOT), between the years 2004 and 2018.


```{css echo=FALSE, message=FALSE, warning=FALSE, CSS}
/* Define a margin before header elements */
h1  {
  margin-top: 52px;
}
h2, h3, h4  {
  margin-top: 48px;
}
/* Define a margin after every first p elements */
p:first-of-type {
  margin-bottom: 14px;
}
``` 

```{r echo=FALSE, message=FALSE, warning=FALSE, Libraries}
# libraries
library(ggplot2)
library(ggpubr)     # ggarrange
library(scales)
library(GGally)     # ggcorr
library(dplyr)
library(tidyverse)
library(knitr)
library(psych)
library(reshape2)  
library(cowplot)
library(gridExtra)

# global settings
knitr::opts_chunk$set(comment=NA, fig.width=9, fig.height=6)

# ggplot layers
mytheme_minimal <- theme_minimal() +
    theme(plot.title = element_text(size = 16, hjust = 0),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 11),
          axis.title.x = element_text(hjust = 0),
          legend.position = "none"
          ) 

mylayer_color <- scale_color_brewer(palette = 'Dark2', direction = 1)  
mylayer_fillcolor <- scale_fill_brewer(palette = 'Dark2', direction = 1)
mylayer_x90 <- theme(axis.text.x = element_text(angle = 90, hjust = 1))
mylayer_hour_breaks <- scale_x_datetime(labels = date_format("%H:%M"), date_breaks = ('2 hours')) 
mylayer_long_xlabel <- theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
mylayer_count_above <- geom_text(stat='count', aes(label = ..count..), vjust = -.2)
```

```{r Load_data}
# Load data
datafile = '../data/collisions_clean.rda'
load(file = datafile)

# Create df_trim_yrs to contain only complete year's worth of data
# TODO: include logic, remove hardcode
df_full_yrs <- subset(df, year != '2019')
```

# Explore

## When

### 1. Collision count by year
```{r}
ggplot(data = df_full_yrs %>% count(year), aes(x = year, y = n)) + 
    geom_bar(stat = "identity", aes(fill = n)) + 
    labs(title = 'Collision count by year', x = '', y = '') +
    mytheme_minimal +
    scale_fill_distiller(palette = 'Greys', direction = 1) + 
    scale_x_discrete(position = 'top') +
    scale_y_continuous(labels = scales::comma) 
```

**Observations: **
There appears to be a 10-is year trend: 5 years down, 5 years up, then going back down.

The first peak, at year 2005, was higher than the second peak 10 years later, at year 2015.

I had expected to see a steady increase in Seattle collisions, mirroring population growth. Perhaps I'm wrong about the steady increase in population, or wrong that the number of **reported** collisions have a strong correlation with the population. #TODO: import population data and compare rate changes with collision rates by year.

### 2. Collision count by hour
```{r warning=FALSE}
ggplot(data = df %>% 
           filter(!is.na(hour)) %>%
           count(hour), 
       aes(x = hour, y = n, group = 1 )) + 
    geom_line() + 
    labs(title = 'Collision Count by Hour', x = '', y = '') +
    mytheme_minimal  +
    scale_y_continuous(labels = scales::comma) 
```

**Observations: **
Collisions tend to rise during commute hours, starting at the lowest point around 4am, reaching a peak at 5pm. 

#### 2a. Hourly collisions by day of week
```{r warning=FALSE}
weekend <- c('Saturday', 'Sunday')
weekday <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
alldays <- c(weekday, weekend)

# TODO fix y max limit programmatically
plot_hours <- function(day, title) {
    gg <- ggplot(data = df %>%
               filter(!is.na(hour) & weekdays(INCDATE) %in% day) %>%
               count(hour),
           aes(x = hour, y = n, group = 1 )) +
        geom_line() +
        labs(title = paste(title), x = '', y = '') +
        mytheme_minimal  +
        scale_x_discrete(breaks = c(0,4,8,12,16,20,24)) +
        scale_y_continuous(limits = c(0, 2250), 
                           breaks = seq(0, 2100, 700),
                           labels = scales::comma) 
    return (gg)
}

# create list of plots, one for each day of the week
plot_list <- lapply(
    as.list(alldays), function(x){
    plot_hours(x, paste0('- ', x, 's'))
})

#cowplot::plot_grid(plotlist = plot_list, ncol = 3)

ggarrange(plotlist = plot_list, ncol=3, nrow=3) + 
    annotate(geom = 'text', label = 'title', hjust = 0)

```


### By month

### By part of day

### By season

## Who

### Striker, strikee

### Pedestrians, Cyclists

### Parked cars, trains, other

## What

### Collision type

### Property damage, Injuries, Fatalities

## Where

### Address type, Junction type

## Why

### Weather, Road Condition, Lighting

### DUI, Distracted, Speeding

### Traffic volume(?)

## Multivariate

### When and Why

### Who and What

### <etc>

# Summary
